\texttt{JAXbind}: Easy bindings to JAX

### 2024-03-12 paper/paper.tex
paper.tex: sync changes

JAX is widely used in machine learning and scientific computing.
Scientific computing relies on existing high-performance code which we
would ideally like to use in JAX. Reimplementing the existing code in
JAX is often impractical and the existing interface in JAX for
connecting custom code requires deep knowledge of JAX and its C++
backend. The aim of \texttt{JAXbind} is to drastically lower the burden
of connecting custom functions implemented in other programming
languages to JAX. Specifically, \texttt{JAXbind} provides an easy-to-use
Python interface for defining custom, so-called JAX primitives
supporting any JAX transformations.

@@

The use of JAX (Bradbury et al., 2018) is widespread in the natural
sciences. JAX's powerful transformation system is of especially high
interest. It enables retrieving arbitrary derivatives of functions,
batch computations, and just-in-time code compilation for additional
performance. Its transformation system relies on all constituents of the
computation being written in JAX.

@@

In this paper, we present \texttt{JAXbind}, a package for bridging any
function to JAX without in-depth knowledge of JAX's transformation
system. The interface is accessible from Python without requiring any
development in C++. The package is able to register any function, its
partial derivatives and their transpose functions as a JAX native call,
a so-called primitive. Derivatives, compilation rules, and batching
rules are automatically registered with JAX.

We believe \texttt{JAXbind} to be highly useful in scientific computing.
We intend to use this package to connect the Hartley transform and the
spherical harmonic transform from ducc (Reinecke, 2024) to NIFTy
(Edenhofer et al., 2024) as well as the radio interferometry response
from ducc with \texttt{resolve} (Arras et al., 2024) for radio
astronomy. Furthermore, we intend to connect the non-uniform FFT from
ducc with JAX for applications in strong-lensing astrophysics. We
envision many further applications within and outside of astrophysics.

@@

Automatic differentiation is a core feature of JAX and often one of the
main reasons for using it. Thus, it is essential that custom functions
registered with JAX support automatic differentiation. In the following,
we will outline which functions our package respectively JAX requires to
enable automatic differentiation. For simplicity, we assume that we want
to connect the nonlinear function \(f(x_1,x_2) = x_1x_2^2\) to JAX. The
\texttt{JAXbind} package expects the Python function for \(f\) to take
three positional arguments. The first argument, \texttt{out}, is a
\texttt{tuple} into which the function results are written. The second
argument is also a \texttt{tuple} containing the input to the function,
in our case, \(x_1\) and \(x_2\). Via \texttt{kwargs\_dump}, potential
keyword arguments given to the later registered Jax primitive can be
forwarded to \texttt{f} in serialized form.

@@

To just-in-time compile the function, JAX needs to abstractly evaluate
the code, i.e.~it needs to be able to know the shape and dtype of the
output of the custom function given only the shape and dtype of the
input. We have to provide these abstract evaluation functions returning
the output shape and dtype given an input shape and dtype for \texttt{f}
as well as for the \texttt{vjp} application. The output shape of the
\texttt{jvp} is identical to the output shape of \texttt{f} itself and
does not need to be specified again.
The abstract evaluation functions take normal positional and keyword arguments.

@@

JAX supports higher order derivatives and can differentiate a
\texttt{jvp} or \texttt{vjp} with respect to the position at which the
Jacobian was taken. Similar to first derivatives, JAX can not
automatically compute higher derivatives of a general function \(f\)
that is not natively implemented in JAX. Higher order derivatives would
again need to be provided by the user. For many algorithms, first
derivatives are sufficient, and higher order derivatives are often not
implemented by the high-performance codes. Therefore, the current
interface of \texttt{JAXbind} is, for simplicity, restricted to first
derivatives. In the future, the interface could be easily expanded if
specific use cases require higher order derivatives.

@@

Currently, \texttt{JAXbind} only has CPU but no GPU support. With some
expertise on Python bindings for GPU kernels adding GPU support should
be fairly simple. We especially want to highlight that the interfacing
with the JAX automatic differentiation engine is identical for CPU and
GPU.

>>>>

JAX is widely used in machine learning and scientific computing, the latter of which often relies on existing high-performance code that we would ideally like to incorporate into JAX.
Reimplementing the existing code in JAX is often impractical and the existing interface in JAX for binding custom code requires deep knowledge of JAX and its C++ backend.
The goal of \texttt{JAXbind} is to drastically reduce the effort required to bind custom functions implemented in other programming languages to JAX.
Specifically, \texttt{JAXbind} provides an easy-to-use Python interface for defining custom so-called JAX primitives that support arbitrary JAX transformations.

@@

The use of JAX (Bradbury et al., 2018) is widespread in the natural
sciences.
Of particular interest is JAX's powerful transformation system.
It enables to retrieve arbitrary derivatives of functions, batch computations, and just-in-time code for additional performance.
Its transformation system relies on all components of the computation being written in JAX.

@@

In this paper, we present \texttt{JAXbind}, a package for bridging any
function to JAX without in-depth knowledge of JAX's transformation
system. The interface is accessible from Python without requiring any
development in C++. The package is able to register any function, its
partial derivatives and their transpose functions as a JAX native call,
a so-called primitive.

We believe \texttt{JAXbind} to be highly useful in scientific computing.
We intend to use this package to connect the Hartley transform and the
spherical harmonic transform from DUCC (Reinecke, 2024) to the probabilistic programming package NIFTy
(Edenhofer et al., 2024) as well as the radio interferometry response
from DUCC with the radio astronomy package \texttt{resolve} (Arras et al., 2024) for radio
astronomy. Furthermore, we intend to connect the non-uniform FFT from
DUCC with JAX for applications in strong-lensing astrophysics. We
envision many further applications within and outside of astrophysics.

@@

Automatic differentiation is a core feature of JAX and often one of the
main reasons for using it. Thus, it is essential that custom functions
registered with JAX support automatic differentiation. In the following,
we will outline which functions our package respectively JAX requires to
enable automatic differentiation. For simplicity, we assume that we want
to connect the nonlinear function \(f(x_1,x_2) = x_1x_2^2\) to JAX. The
\texttt{JAXbind} package expects the Python function for \(f\) to take
three positional arguments. The first argument, \texttt{out}, is a
\texttt{tuple} into which the results are written. The second
argument is also a \texttt{tuple} containing the input to the function,
in our case, \(x_1\) and \(x_2\). Via \texttt{kwargs\_dump}, any
keyword arguments given to the registered JAX primitive can be
forwarded to $f$ in a serialized form.

@@

To just-in-time compile the function, JAX needs to abstractly evaluate
the code, i.e.~it needs to be able to infer the shape and dtype of the
output of the function given only the shape and dtype of the
input. We have to provide these abstract evaluation functions returning
the output shape and dtype given an input shape and dtype for $f$
as well as for the \texttt{vjp} application. The output shape of the
\texttt{jvp} is identical to the output shape of $f$ itself and
does not need to be specified again.
The abstract evaluation functions take normal positional and keyword arguments.

@@

JAX supports higher order derivatives and can differentiate a
\texttt{jvp} or \texttt{vjp} with respect to the position at which the
Jacobian was taken. Similar to first derivatives, JAX can not
automatically compute higher derivatives of a general function \(f\)
that is not natively implemented in JAX. Higher order derivatives would
again need to be provided by the user. For many algorithms, first
derivatives are sufficient, and higher order derivatives are often not
implemented by high-performance codes. Therefore, the current
interface of \texttt{JAXbind} is, for simplicity, restricted to first
derivatives. In the future, the interface could be easily expanded if
specific use cases require higher order derivatives.

@@

Currently, \texttt{JAXbind} only supports primitives that act on CPU memory.
In the future, GPU support could be added analogous to the CPU backend.
Note, the automatic differentiation in JAX is backend agnostic and would thus not require any additional bindings.
